---
title: "Soapcheckr"
output:
  md_document:
    variant: gfm
    fig_width: 10
execute:
   eval: FALSE
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures"
)
```

# soapcheckr

## Check whether a soap film smoother boundary and knots make sense

Setting up a soap film smoother is often a hard and frustrating process. This packages provides functions that checks the boundary, knots, and data that you feed to a soap are "correct". The functions in the package will also plots the area that will be modeled by the soap film smoother in red, and knots and/or data as circles as viable locations. Any circles that have a X through them are considered inadequte and will need to be removed.

To do the spatial analysis, we require a couple of extra libraries: [{sf}](https://r-spatial.github.io/sf/) and [{mgcv}](https://cran.r-project.org/web/packages/mgcv/index.html). That will be loaded when installing [{soapcheckr}](https://github.com/dill/soap_checker). 

`{soapcheckr}` can be installed using the following code; 

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dill/soap_checker")
```

Next load `{soapcheckr}` and walk through the vignette to see how to use the functions. We highly encourage walking through both examples in the vignette. 

```{r, eval = FALSE}
library(soapcheckr)
vignette("how_to_use_soapcheckr")
```

## Other tips

* Note that you need to set the `k` and `nmax` argument in `autocruncher()` to be the same as your planned value in `gam()` for your soap film smoother
* The functions assumes that your spatial variable names are `x` and `y` unless you supply the naming arguments with the column names from your data.
* Make sure that your locations are in Northings/Eastings, (e.g., UTMs). Using latitude and longitude (e.g., decimal degree) will give strange results (as the soap film smoother is isotropic so treats 1 unit change in either dimension is equal, this isn't true for lat/long!).
* Sometimes you need to increase the tolerance (e.g `tol = 1e-6`) argument when using `soap_check()`
* Note that the boundary must be a `list` of `list`s or `data.frame`s. So if you have a polygon boundary with your boundary vertices in it, that must be wrapped in a `list`! The vector names of each boundary need to be names `x` and `y` for the functions to work. 


Written by David L. Miller and Benjamin L. Hlina. Released under the GPL (version 2).
