---
title: "How to use Soapcheckr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use Soapcheckr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.dim = c(7, 5), 
  comment = "#>"
)
```

### Our Objectives

The purpose of this vignette is to demonstrate an effective workflow while using [{soapcheckr}](https://github.com/dill/soap_checker) to efficiently make a soap-film smoother for a Generalized Additive Model (GAM) using the package [{mgcv}](https://cran.r-project.org/web/packages/mgcv/index.html). Soap-film smoothers are really useful when trying to model a variable within a 3-dimensional space (e.g., bathymetry of a lake; [Gavin Simpson Blog Post](https://fromthebottomoftheheap.net/2016/03/27/soap-film-smoothers/)). They can be used for all sorts of data but are quite complex and difficult to setup. [{soapcheckr}](https://github.com/dill/soap_checker) tries to make this process a little easier. This vignette will walk through several different examples using example data sets loaded when loading [{soapcheckr}](https://github.com/dill/soap_checker) and [{mgcv}](https://cran.r-project.org/web/packages/mgcv/index.html). 

### Example 1: Making a soap-film smoother for a Simple boundary - Ramsay's horseshoe 

If we were wanting to make a soap-film smoother for a boundary that does not have any inner boundaries (e.g., a lake without an island) we can use `{soapcheckr}` to assess our single boundary and the potential knots we want to smooth over. 

#### Step 1: Load packages 

We will use some data manipulation functions from [{dplyr}](https://dplyr.tidyverse.org/) and some spatial functions from [{sf}](https://r-spatial.github.io/sf/) to first check if we can make a soap-film using [{soapcheckr}](https://github.com/dill/soap_checker)) and second to run a GAM using [{mgcv}](https://cran.r-project.org/web/packages/mgcv/index.html) with a soap-film smoother.

```{r setup, message = FALSE}
{
  library(broom.mixed)
  library(fitdistrplus)
  library(dplyr)
  library(ggplot2)
  library(gratia)
  library(mgcv)
  library(purrr)
  library(soapcheckr)
  library(sf)
}
```

#### Step 2: Check if we can make a soap-film for Ramsey's horseshoe

We will load Ramsey's horseshoe from [{mgcv}](https://cran.r-project.org/web/packages/mgcv/index.html) and make it a list within a list to sufficiently create the boundary a soap-film smoother needs within a GAM. 

```{r ramsays horseshoe}
fsb <- list(fs.boundary())
```

We can then can check the boundary using `soap_check()`
```{r check ramsays horseshoe}
soap_check(fsb)
```

We can see that `soap_check()` returns `TRUE` indicating that we can use this boundary to make a soap-film smoother. `soap_check()` will assess if the boundary supplied has any overlapping polygons and in the correct structure for a soap-film smoother to be run in `{mgcv}`. 

#### Step 3: Check if evenly spaced knots and data will work for soap-film smoother

Sometimes knots are too close to the boundary, resulting with errors from the model that look like:

```
Error in crunch.knots(ret$G, knots, x0, y0, dx, dy) :
knot 1 is on or outside boundary
```
It can be tedious and annoying to try to figure out which knots and/or data points are causing issues. The below work flow will demonstrate using an example knots and dataset of how to effectively remove knots and/or data points that result in the above error within the GAM. 

We will use `expand.grid()` to create an equally spaced grid of knots. Then we remove knots outside of the boundary using `inSide()` from [{mgcv}](). There still may be knots that are too close to the boundary that will cause for a soap-film smoother to not work. 

```{r build fsb knots, message = FALSE}
# create knots 
knots <- expand.grid(x = seq(min(fsb[[1]]$x), 
                             max(fsb[[1]]$x), len = 15),
                     y = seq(min(fsb[[1]]$y) + 0.05,
                             max(fsb[[1]]$y), len = 10))
x <- knots$x
y <- knots$y

# identify the knots that are outside the boundary 
ind <- inSide(fsb, x = x, y = y)
# remove knots outside the boundary 
knots <- knots[ind, ]
```

We will also create some fake data to test the model. We will use a uniform distribution to make our test data and a response variable will be created using `fs.test()`. We will use `inSide()` to remove data that fall outside our boundary 

```{r build fake data, message = FALSE}
set.seed(0)
n <- 600

# Our x and y data 
x <- runif(n) * 5 - 1
y <- runif(n) * 2 - 1

# create our response variable 
z <- fs.test(x, y, b = 1)

## remove outsiders

ind <- inSide(fsb, x = x, y = y) 

z <- z + rnorm(n) * 0.3 ## add noise

# create the data we want to model 
dat <- data.frame(z = z[ind],
                  x = x[ind],
                  y = y[ind])
```

However, there may still be knots and/or data points that are too close to the boundary.
One can go through, one-by-one and remove the offending knots as `crunch.knots()` finds them, but that’s a bit tedious. Enter `soap_check()` and `autocruncher()`, with the former allowing one to visually check what knots and/or data will causes issues and the latter identifying the index location of the offending knots or data points. If the knot dataframe has columns names other than `x` and `y`, we need to supply `soapcheckr()` those names using the arguments `x_name` and `y_name`, respectfully. 
```{r check our knots}
soap_check(fsb, knots = knots)
```
We can see that there four offending knots that we can subsequently remove using `autocruncher()`. The function will return the indices of knots that would cause issues. If the  knot dataframe has columns names other than `x` and `y`, we need supply `autocruncher()` those names using the arguments `xname` and `yname`, respectfully. Note that you need to set the `k` argument in `autocruncher()` to be the same as your planned value in `gam()` and if you set the `nmax()` parameter for the soap film you need to provide that to `autocruncher()` too.
```{r remove knots that are offending}
crunch_index <- autocruncher(fsb, knots, k = 30)
crunch_index

# remove knots that are problematic
knots <- knots[-crunch_index, ] 
```

We can use `soap_check()` again to check if knots all fall within the boundary. 
```{r}
soap_check(fsb, knots = knots)
```

And they do! Congratulations! 

### Step 4: Check the data 

We can also use `soap_check()` to check if our data falls within the boundary, but `soap_check()` only cares about the coordinates you want to supply the soap-film. So first we will create a secondary dataframe that has the response variable, `z`, removed from it. 

```{r}
dat_2 <- dat[, 2:3]
soap_check(fsb, data = dat_2)
```

Congrats we can see our data doesn't fall too closely to boundary and is within the boundary. 

### Step 5: Run the model 

Between data, knots, and boundary all column names must be the same for the model to work. Prior to running confirm that they are all the same. 

```{r run GAM}
m <- gam(z ~ s(x, y, k = 30 , bs = "so",
               xt = list(bnd = fsb)),
         knots = knots, 
         data = dat)

```
Next, check main effects of the model. 
```{r check main effects of model, eval = FALSE}
anova(m)
```
Then check the model effects using `draw()` from [{gratia}](https://gavinsimpson.github.io/gratia/)
```{r draw model effects}
draw(m)
```
Lastly, check the model fit using `appraise()` from [{gratia}](https://gavinsimpson.github.io/gratia/)
```{r check model fit}
appraise(m)
```
We can see the model fits well and is appropriate for the example data. Now that you've walked through a simple boundary, we will go to a more complex boundary example. More than likely you will be working with a complex boundary. The below walk through can applied to a simple boundary as well. 

### Example 2: Making a soap-film smoother for a more complex boundary.
More than likely you will have your boundary as a `sf` object. To convert that `sf` object into the boundary list needed by {mgcv} and {soapcheckr}, we will have to do some conversions. First that boundary might not be in the correct coordinate reference system (CRS). To create a soap-film smoother we need to use a CRS that for one unit change in either dimension (i.e., x and y) are equal. For example using latitude and longitude in decimal degrees with a WGS84 project will not work because one unit of change in either direction is not equal. Therefore, we will need to use a CRS that is based on an equal units. The most common CRS to do this is UTMs, if the boundary is already in UTMs great, if not see below 

#### Step 1: Convert CRS 
The more complex boundary that we will load through {soapcheckr} is a lake from northern Wisconsin. I grew up fishing on this lake that intially got me interested in aquatic ecosystems and fish. This lake falls within UTM zone 15N that can also be refereed to as ESPG: 32615, but your boundary will more than likely fall into a different CRS. You can look up ESPG codes [here](https://epsg.io/).  

```{r sissabagama lake crs, message = FALSE}
sissabagama_lake_sf <- sissabagama_lake_sf %>% 
  st_transform(crs = 32615)
```

#### Step 2: Convert to boundary list  
We need to create the list of lists of the boundaries from the `sf` object that we will supply othe soap-film smoother.  

Our example lake has a `geometry` column that is a `POLYGON`. We need to be able to split that into each polygon (i.e., islands) that we will create the boundary list from. We can do this by first casting our `geometry` into `MULTIPOINT` and assigning each `MULTIPOINT` row an ID value. 
```{r sissabagama lake pnt sf, message = FALSE}
bnd_pt_sf <- sissabagama_lake_sf %>%
  dplyr::select(geometry) %>%
  st_cast("MULTIPOINT") %>%
  mutate(
    id = 1:nrow(.)
  )
```

Next we will split our `sf` object and iterate over each `MULTIPOINT` geometry to first cast to individual `POINT` geometry and extract each `x` and `y` coordinates. It is important in this step that the names of the coordinates are `x` and `y`.

```{r sissabagama lake pnt, message = FALSE, warning = FALSE}
bnd_pt <- bnd_pt_sf %>%
  split(.$id) %>%
  purrr::map(~ st_cast(.x, "POINT") %>%
               mutate(
                 x = st_coordinates(.)[,"X"],
                 y = st_coordinates(.)[,"Y"]
               ) %>%
               st_drop_geometry() %>% 
               dplyr::select(-id)
  )
```

We now have a list of dataframes split by  each polygon's x and y coordinates that have had the `id` column removed. We then need to create a vector that is `id` number of each polygon. In this case its 1-5. You will need to change the 5 to whatever number of polygons your boundary has (e.g., 20).

We will then iterate over our list and bind them all into together to get our lists and lists of our polygon boundaries. 

```{r create list of lists boundary, message = FALSE}
nr <- 1:length(bnd_pt)

sissabagama_bnd_ls <- lapply(nr, function(n) as.list.data.frame(bnd_pt[[n]]))
```

### Step 3: Check if the boundary list works 

We will check the boundary list using `soap_check()`

```{r check more complex boundary}
soap_check(sissabagama_bnd_ls)
```
`soap_check` returns back `TRUE` so our more complex boundary will work for our soap-film smoother.

#### Step 4: Make knots for a more complex boundary using [{sf}]()

We can use `st_make_grid()` to create a grid of equally spaced points across the boundary box of our example `sf` object of sissabagama lake. 

```{r make knot grid, message = FALSE}
lake_grid <- sissabagama_lake_sf %>%
  st_make_grid(cellsize = 200, square = TRUE, what = "centers") %>%
  st_as_sf() 

st_geometry(lake_grid) <- "geometry"
```

We then will remove all the knots that fall outside the boundary by using `st_intersection()`. 

```{r remove all knots that are outside boundary, message = FALSE}
lake_intesects <- st_intersection(sissabagama_lake_sf, lake_grid)
```

Next we will create our knot dataframe by extracting the lon and lat of each point and then dropping the `geometry column` and selecting our `lon` and `lat` columns. 

```{r make knot dataframe, message = FALSE}
lake_knots <- lake_intesects %>%
  mutate(
    lon = st_coordinates(.)[,"X"],
    lat = st_coordinates(.)[,"Y"]
  ) %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(lon, lat)
```

Now that we have our knots we can check to see if there are any knots that fall too close to the boundary using `soap_check()`
```{r check knots using soap_check}
soap_check(sissabagama_bnd_ls, knots = lake_knots, 
           x_name = "lon", y_name = "lat")
```


We can see that there are a few knots that are too close to the boundary. We can remove them using `autocruncher()`
```{r}
nrow(lake_knots)
crunch_ind <- autocruncher(sissabagama_bnd_ls, lake_knots, k = 37, 
                           xname = "lon", yname = "lat")
crunch_ind

# remove knots that are problematic
lake_knots <- lake_knots[-crunch_ind, ] 
```
Now that those knots have been removed we can recheck our knots using `soap_check()`
```{r recheck knots}
soap_check(sissabagama_bnd_ls, knots = lake_knots, 
           x_name = "lon", y_name = "lat")

```
Congratulations! We have knots and a boundary that we can supply our model. 

### Step 5: Sissabagama example dataset 

We will bring our sampled depth at given locations for Sissabagama lake. This data was generated by referencing the contour map supplied by the [Wisconsin DNR](https://apps.dnr.wi.gov/lakes/lakepages/LakeDetail.aspx?wbic=2393500). We will first remove depth to check if the data points all fall within our boundary.  


```{r remove depth for soap_check}
sissabagama_bath_pt <- sissabagama_bath %>% 
  dplyr::select(-depth)

soap_check(sissabagama_bnd_ls, data = sissabagama_bath_pt)
```
Then we will assess the data to see what distribution the data likely will fit using functions from [{fitdistrplus}](https://cran.r-project.org/web/packages/fitdistrplus/index.html). 
```{r check skewness and kurtosis}
depths <- sissabagama_bath$depth
descdist(depths)
```

We can see that skewness and kurtosis indicate that a Gamma distribution will likely fit. 
```{r check distr}
fit_gamma <- fitdist(depths, distr = "gamma", method = "mme")
plot(fit_gamma)
```

We can see the data appears that depth data will likely  fit a gamma distribution.

### Step 6: Run the model 
Prior to running our GAM with a soap film smoother we need to add one last thing to our boundary list. We need to add the variable `f` to every boundary polygon within our boundary list. This variable indicates to the soap-film smoother that our response variable is `0` right at the boundary. 
```{r add f to our boundary list}
names(lake_knots) <- c("x", "y")

sissabagama_bnd_ls <- lapply(nr,
                             function(n)
                               sissabagama_bnd_ls[[n]] <- c(
                                 sissabagama_bnd_ls[[n]],
                                 list(f = rep(0, 
                                              length(sissabagama_bnd_ls[[n]]$x)
                                 )
                                 )
                               )
)
```

 
We can now successfully run our GAM with a soap-film smoother.

```{r model our depth}
m1 <- gam(depth ~ s(x, y,
                bs = "so",
                xt = list(bnd = sissabagama_bnd_ls)),
          family = Gamma(link = "log"),
          knots = lake_knots,
          data = sissabagama_bath)
```

We can evaluate the partial effects of the model using `draw()` from `{gratia}`
```{r check partial effects}
draw(m1)
```
We can evaluate how well the model fit is to the data using `appraise()` also from `{gratia}`
```{r check model fit for m1}
appraise(m1)
```
### Step 7: Plot our predicted results. 
We will first create a 10 m grid from our `sf` object of our boundary. Depending on the size of the boundary you can change the grid size distance to whatever distance makes sense (i.e., if the system is large you may want to increase the `cellsize`). 
```{r create 10m grid to efficiently predict over the boundary}
lake_pred <- sissabagama_lake_sf %>%
  st_make_grid(cellsize = 10, square = TRUE, what = "centers") %>% 
  st_as_sf() 
st_geometry(lake_pred) <- "geometry"
```
After creating the grid that we will predict on we will need to remove any points that fall outside our polygon bundary. 
```{r remove all points that fall outside the boundary, warning = FALSE}
lake_pred <- st_intersection(lake_pred, sissabagama_lake_sf) %>% 
  dplyr::select(geometry)
```
Then we extract latitude and longitude and convert the `sf` object into a dataframe. 

```{r convert to a dataframe}
lake_pred_df <- lake_pred %>% 
  mutate(
    x = st_coordinates(.)[,"X"], 
    y = st_coordinates(.)[,"Y"], 
  ) %>% 
  st_drop_geometry()

```
We then can use the function `augment()` from the package [{broom.mixxed}](https://cran.r-project.org/web/packages/broom.mixed/index.html) to predict depth of the lake at a given latitude and longitude. We will also need to back transform the fitted data using the e. 

```{r predict using augment from broom.mixxed}
pred <- augment(m1, newdata = lake_pred_df)
pred <- pred %>% 
  mutate(
    lower = exp(1) ^ (.fitted - 1.96 * .se.fit),
    higher = exp(1) ^ (.fitted + 1.96 * .se.fit),
    .fitted = exp(1) ^ .fitted
  )
glimpse(pred)
```

Lastly, we can visualize our predicted depths to create a bathymetic map of the lake using `ggplot()`. 

```{r plot our predicted depths using ggplot}
ggplot() +
  geom_raster(data = pred, aes(x = x, y = y, fill = .fitted)) +
  geom_sf(data = sissabagama_lake_sf, fill = NA, colour = "black") +
  scale_fill_viridis_c(name = "Depth (m)",
                       trans = "reverse",
                       breaks = rev(seq(0, 60, 15))
  ) + 
  theme_void(
    base_size = 15
  ) + 
  theme(
    legend.background = element_blank(),
    legend.position = c(0.98, 0.82),
  ) + 
  guides(fill = guide_colourbar(
    frame.linewidth = 0.3,
    ticks.colour = 'black', 
    frame.colour = 'black')) + 
  labs(x = "Longitude", 
       y = "Latitude")

```
